### 2018-05-14 4_impl.tex
write interpreter. update pictures.

Each memory unit has an \texttt{XType} associated with it.
The X-type is a symbolic representation of the C primitive type%
%
\footnote{Here we should note that \porthos[2] can eventually evolve to be able to analyse programs written in an OOP language (for instance, in C++). In this case, the \texttt{XType} will have more complex structure than a simple enumeration, which it has when we need to emulate only primitive types of C language. See more detailed discussion on input language type system in Section~\ref{ch:impl:proc:x-pre-compiler:type}.} %TODO!!!!! Check that XType is really a simple enumeration
%
that is easily convertible to an SMT-type (modelled as \texttt{ZType}).%TODO: add the class ZType

@@

  \1 \texttt{XFakeEvent} \\
    The fake events are the auxiliary elements of \xgraph{}.
    %\begin{itemize}[noitemsep]
    \2 \texttt{XEntryEvent}, the per-process unique source event in the event-flow graph,
    \2 \texttt{XExitEvent}, the process sink event. There are two kinds of sink events: \textit{complete} and \textit{incomplete} (see description in Section~\ref{ch:impl:proc:x-unroll}),
    \2 \texttt{XNopEvent}, the no-operation event (a jump to the next event), used for correct encoding in case when the control-flow branch does not have any event (see Figure~\ref{encode:branching:nop}).
   % \end{itemize}
%\end{itemize}
\end{outline}

@@

<TODO: in picture change: xmemorymaager belongs to the program interpreter!>

@@

<TODO: ADD XTypeManager to the picture!>

We distinguish three types of processes (each implement the \texttt{XInterpreter} interface):
\begin{itemize}[noitemsep,topsep=0pt]
\item the prelude process \texttt{XPreludeInterpreter} (to be executed before all other processes) that allows only declaration of local or shared variables, computations and memory operations;
\item the postlude process \texttt{XPostludeInterpreter} (to be executed after all other processes) that allows only declaration of local variables, memory operations, computations and assertions;
\item the process \texttt{XProcessInterpreter} (to be executed in parallel) that allows all X-compiler interface operations except declaration of shared variables %(that should be declared during pre-compilation stage or during compilation of the prelude process)
and program assertions (declarations of local variables, computations, barriers, unconditional jumps, non-linear branching statements, method calls).
\end{itemize}

@@

The \xgraph{} abstract machine model disposes infinitely many memory units, both local and shared (see Section~\ref{ch:impl:model:xgraph:mem}).
The X-compiler accesses all memory units via the \texttt{XMemoryManager}, which by the end of pre-compilation stage is already initialised (has registered all shared memory units).
However, the memory manager is a stateful component of the compiler as it offers the methods for declaring and removing local memory units dynamically at the compilation time.
At each time of the compilation process, the \texttt{XMemoryManager} can resolve the memory unit by its name.
Following the C standard, local memory units have higher priority over global ones.
Since C language allows usage of variables that have the same name to be declared in nested contexts, the \texttt{XMemoryManager} carries the stack of block block-contexts for local variables. %TODO: refactor XMemoryManager!

<some picture ? >

@@

Current version of \porthos[2] contains two invocation hooks, the \texttt{XPorthosInvocationHook} for describing the compatibility-mode functions of the input \porthos[2] input language, and \texttt{XKernelInvocationHook} for describing the Linux kernel-specific functions.
An invocation hook can model any type of operations (memory, fence, computational, etc.).
For instance, the \texttt{XPorthosInvocationHook} intercepts calls \texttt{`location.store(atomic, register)'} and processes the store operation with respect to the specified atomic.
%TODO: more about described functions
%TODO: rename InvocationHook -> XInvocationHook

@@

stateful

X compilation error

contexts

computation: how not to collect (in which case we loose)

stacks: context, readyContext, call, 
  BlockContext (rename): what stores, enum of states

almostReadyContexts: for computation events. Flushed by invocation . When are filled.

setting up edges depending of type of ready-context

- hierarchy of Compilers: pic on inheritance (XCompiler is an stateful abstract machine). Restrictions of lude-compilers
  postlude: assertion: a new event XAssertionEvent

interfaces that offer compilers: listing of interface

algorithm for interpreting

@@

up to bound $k$

recursion -- may be, if calls are supported

@@

Z-type : SMT-specific (say about smt-lib)

@@

\subsubsection{SMT-formula translator (11)}
\label{ch:impl:proc:smt-solver}%TODO: ref it in pre-word in this chapter
%TODO: check that we have ALL references in pre-word in this chapter

%todo: mention The SMT-LIB standard http://smtlib.cs.uiowa.edu/papers/smt-lib-reference-v2.6-r2017-07-18.pdf

@@

structure: verdict

%TODO: say sth about syntax errors handling: still now, what if parser does not accept syntax?

@@

>>>>

Each memory unit has an \texttt{XType} associated with it.
The X-type is a symbolic representation of the C primitive type%
%
\footnote{Here we should note that \porthos[2] can eventually evolve to be able to analyse programs written in an OOP language (for instance, in C++). In this case, the \texttt{XType} will have more complex structure than a simple enumeration, which it has when we need to emulate only primitive types of C language. See more detailed discussion on input language type system in Section~\ref{ch:impl:proc:x-pre-compiler:type}.} %TODO!!!!! Check that XType is really a simple enumeration
%
that is easily convertible to an SMT-type (modelled as \texttt{ZType}).%TODO: add the class ZType
As the type of an X-memory unit may not be resolved correctly, the memory units keep this information as a boolean flag.

@@

  \1 \texttt{XFakeEvent} \\
    The fake events are the auxiliary elements of \xgraph{}.
    %\begin{itemize}[noitemsep]
    \2 \texttt{XEntryEvent}, the per-process unique source event in the event-flow graph,
    \2 \texttt{XExitEvent}, the process sink event. There are two kinds of sink events: \textit{complete} and \textit{incomplete} (see description in Section~\ref{ch:impl:proc:x-unroll}),
    \2 \texttt{XNopEvent}, the no-operation event (a jump to the next event), used for correct encoding in case when the control-flow branch does not have any event (see Figure~\ref{encode:branching:nop}),
    \2 \texttt{XAssertionEvent}, the reachability assertion made at the postlude statement of the program.
    An assertion is modelled as an event for the purpose of encoding the postlude statement as a separate process that is compatible with the encoder.
   % \end{itemize}
%\end{itemize}
\end{outline}

@@

@@

We distinguish three types of processes (each implementing the \texttt{XInterpreter} interface):
\begin{itemize}[noitemsep,topsep=0pt]
\item the prelude process \texttt{XPreludeInterpreter} (to be executed before all other processes) that allows only declaration of local or shared variables, computations and memory operations;
\item the postlude process \texttt{XPostludeInterpreter} (to be executed after all other processes) that allows only declaration of local variables, memory operations, computations and assertions;
\item the process \texttt{XProcessInterpreter} (to be executed in parallel) that allows all X-compiler interface operations except declaration of shared variables %(that should be declared during pre-compilation stage or during compilation of the prelude process)
and program assertions (i.e., it allows declarations of local variables, computations, barriers, unconditional jumps, non-linear branching statements, and method calls).
\end{itemize}

@@

The \xgraph{} abstract machine model disposes infinitely many memory units, both local and shared (see Section~\ref{ch:impl:model:xgraph:mem}).
The X-compiler accesses all memory units via the \texttt{XMemoryManager}, which by the end of pre-compilation stage is already initialised (has registered all shared memory units).
However, the memory manager is a stateful component of the compiler as it offers the methods for declaring and removing local memory units dynamically at the compilation time.

The interface methods exposed by the \texttt{XMemoryManager} are presented in Figure~\ref{fig:mem-manager}.
At each time of the compilation process, the \texttt{XMemoryManager} can resolve the memory unit by its name.
Following the C standard, local memory units have higher priority over global ones (the method \texttt{getDeclaredUnitOrNull} returns the first memory unit found, either a global one or a local one, or null if no memory units with requested name have been registered).
Since C language allows usage of variables that have the same name to be declared in nested contexts, the \texttt{XMemoryManager} carries the stack of block contexts for local variables. %TODO: refactor XMemoryManager!

\begin{figure}[h]
\centering
\lstinputlisting[language=java]{inc/lst/XMemoryManager.java}
\caption{X-memory manager public interface}
\label{fig:mem-manager}
\end{figure}

@@

Current version of \porthos[2] contains two invocation hooks, the \texttt{XPorthosInvocationHook} for describing the compatibility-mode functions of the input \porthos[2] input language, and \texttt{XKernelInvocationHook} for describing the Linux kernel-specific functions.
An invocation hook can model any type of operations (memory, fence, computational, etc.).
For instance, the \texttt{XPorthosInvocationHook} intercepts calls \texttt{`location.store(atomic,register)'} and processes the store operation with respect to the specified atomic.
%TODO: more about described functions
%TODO: rename InvocationHook -> XInvocationHook

@@

%In this paragraph we discuss only the implementation of the process interpreter \texttt{XProcessInterpreter}.
%This interpreter is used to build the branched event-flow graph with all kinds of jumps, memory operations and fences.
The X-program interpreter is invoked by the \texttt{Y2XConverterVisitor} which walks down the recursive syntax tree \ytree{}.
The calls to the X-program interpreter are dispatched to currently maintained X-process interpreter.
To be able to properly recognise nested statements of the \ytree{}, the X-process interpreter needs to have stacks.
On any semantic error, the X-interpreter throws an \texttt{XInterpretationError}.

The interface methods of the \texttt{XInterpreter} are presented in Appendix~\ref{apx:xinterpreter}.
Note the clear modular independence of the X-interpreter as it has no methods that operate entities of the Y-level (all conversion work with Y-entities is done by the \texttt{Y2XConverterVisitor}).

All interface methods can be divided into two groups.
The first group constitute methods that emit events.
These methods construct an event object (as events contain the \texttt{XEventInfo} structure that stores information about the owning process, all events must be created by the process constructor, i.e. X-interpreter) and change the state of interpreter considering the newly created event.
Note that the methods \texttt{createComputationEvent} do not emit a computation event but only create one.
This is performed as optimisation that removes unused computations from the model (otherwise, for instance, the assignment 
`\texttt{x = 1 * (2 + 3)}' would be compiled to two consequent events `\texttt{eval(2 + 3); write(register <- 1 * (2 + 3));}' instead of single event `\texttt{write(register <- 1 * (2 + 3));}').
If the computation event was not used at all (for example, in the following C code: `\texttt{foo(); 1; bar();}' the execution event `\texttt{eval(1)}' is not used by the model), it is also removed from the event-graph (see justification in Section~\ref{ch:impl:model:xgraph:evt}).
The second group of X-interpreter methods consists of the methods for defining non-linear statements (branchings and loops).
These methods change the state of the interpreter and set up additional non-linear control-flow edges.
%`\texttt{register = 1 * (location + 2)}' would be compiled to three consequent events `\texttt{load(temp\_reg\_1 <- location); eval(temp\_reg\_1 + 2); write(register <- 1 * (temp\_reg\_1 + 2));}',

As a high-level instruction (Y-level) may be compiled into a sequence of low-level instructions (for example, a computation that involves shared variables should firstly load them into the local memory and then process the computation event over local-only memory), the interpreter must maintain the \textit{stack of contexts} and remember the \textit{previous event} to be able to correctly process nested non-linear statements of C language.
The context is a data structure that carries the \textit{state} and some additional information in case of processing non-linear statements (e.g., conditional event, first and last then- and else-branch events for binding, etc.).
Once the new event has been emitted, the interpreter sets up control-flow edges w.r.t. state of the context on the top of the stack (the context stack is always non empty: the bottom context is a linear one).
The context state is an enumeration of the following values:
\begin{itemize}
\item \texttt{WaitingAdditionalCommand}: the interpreter is in the state of defining the complex (non-linear) statement and is not able to process any new event (will throw an exception if any) until the state is not changed,
\item \texttt{WaitingFirstSubBlockEvent}, the first next emitted event will be accepted as the first event of the branching statement (later the interpreter will set up jumps to that event),
\item \texttt{WaitingNextLinearEvent}, the standard interpreter state for processing next linear event,
\item \texttt{Idle}, the interpreter does not set up the edge from the previous event to the new event.
\end{itemize}

Once the (non-linear) context has been popped out of the context stack (by the methods of the second group that finish non-linear branch definition) and added to the queue of \texttt{almostReadyContexts}.
The almost-ready-context becomes a ready-context (i.e., it moves to the queue \texttt{readyContexts}) once the non-linear statement definition has finished (the method \texttt{finishNonlinearBlockDefinition}).

Before processing a newly emitted event, the interpreter checks whether the queue \texttt{readyContexts} is not empty.
If so, it iterates over all ready-contexts and sets up control-flow edges considering that the newly emitted event is an exit-event" of the non-linear context (e.g., for a branching context the interpreter adds primary edges from condition-event to the first-true-branch-event and from the last-true-branch-event to the exit-event, the same is done with alternative edges for the false-branch).

%The \texttt{almostReadyContexts} queue is necessary for not adding erroneous edges on

For simplicity of the interpretation, we keep the jump events (no-computation events) in the flow-graph, however they can be removed and the edges to and from them can be rewritten if necessary.

%algorithm for interpreting

%The method \texttt{tryConvertToLocalOrNull} makes an attempt to convert an arbitrary X-entity to the local memory unit: registers and computation events are casted and returned by the method as-is, shared memory units are copied to a temporary register (\texttt{declareTempRegister}), which is returned by the method.
%In other cases, the method returns \texttt{null}.

@@

The compiled graph needs to be unrolled up to the user-defined bound $k$.
Comparing to \porthos[1], we have change the meaning of the bound: instead of executing all cycles at most $k$ times, the \porthos[2] interprets the bound as the maximum number of Y-instructions to be executed after the unrolling.
The meaning of a bound was changed in order to increase predictability of the size of the result SMT-formula.

Note, recursive function calls create unconditional jumps back in the event-flow graph, which will also be recognised as a loop and be affected by the unrolling algorithm.

@@

@@

As the \zformula{} follows the SMT-LIB standard, its translation to an SMT-formula constitutes almost one-to-one mapping.
The translation is performed by the stateless visitor \texttt{Z2SmtTranslator}.

\subsubsection{SMT solver (11)}
\label{ch:impl:proc:smt-solver}
%TODO: check that we have ALL references in pre-word in this chapter
As it has been mentioned in Section~\ref{ch:impl:principles}, \porthos[2] uses the third-party SMT-solver \tool{Z3}, which offers the Java API for constructing requests to the Z3 solver and interpreting its responses.

@@

The result of execution of \porthos[2] is the verdict modelled by the class \texttt{AppVerdict}.
This is a structure that contains the result of analysis and auxiliary information such as collected errors and time of execution.
The app verdict may be rendered to any format convenient for the user.

@@

There are some ...
